---
title: "Project EDA"
author: "Karan Thacker"
date: "28 August 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Introdution 
This is a project on doing Exploratory Data Analysis on a set of observations of white wine. 
**Objective** to be acheived is to get an idea about which variables are closely related to the output variable `quality` of the 
data set. To acheive this objective, the primary tool is the `ggplot` package to create plots including : 

- scatter plots 
- histograms 
- box plots 
- bar charts 
- line graphs 

The procedure implemented is as follows:

1. Examining the data structure 
2. exploring single variable
3. exploring 2 variable relationship
4. exploring multi variable relationship


##  Examining the data structure 

below is the structure of the data frame of white wine which is named `ww` and the type of each variable is shown. Also the stats of each
variable is given by the summary function

```{r  echo=FALSE, message =FALSE, warning =FALSE}
ww = read.csv("wineQualityWhites.csv") # function to write csv data to data frame
str(ww)     # to see the data type of each variable 
summary(ww) # summarises all the varaibles with stats
```

```{r  echo=FALSE, message =FALSE, warning =FALSE}
# installing the libraries for the functions used with library() function
library(ggplot2)
library(gridExtra) 
library(dplyr)
library(reshape2)
library(GGally)
```


## exploring single variable

summarising the output variable quality with stats of mean median max min etc
```{r  echo=FALSE, message =FALSE, warning =FALSE}
summary(ww$quality)
table(ww$quality)
```
we can see that **9** is the top score of `rating` variable with median score of **6**

```{r echo=FALSE, message =FALSE, warning =FALSE}
# creating function for plotting histogram
create_hist = function(xvariable, xname ){
  return(qplot(x = xvariable, data = ww, xlab = xname ))
  # qplot is used to plot graphs and as.factor used 
  #to change the variable from int to factor 

}

```


### plotting histogram for quality 

```{r  echo=FALSE, message =FALSE, warning =FALSE}
create_hist(as.factor(ww$quality),xname = 'quality')
```

### Categorising the output variable
```{r  echo=FALSE, message =FALSE, warning =FALSE}
# using if else statement to create a quality group out of ratings
ww$quality_G = ifelse(ww$quality<= 5,'average',
                      ifelse(ww$quality<=6,'good','classic'))
ww$quality_G = ordered(ww$quality_G,levels = c('average','good','classic'))  
create_hist(ww$quality_G, xname = 'quality group') +geom_bar()
```
```{r echo=FALSE, message =FALSE, warning =FALSE}
table(ww$quality_G)
```

There is a new variable introduced which is `quality_G' which is categorised into 3 groups :

1. rating (0-5) -> average
2. rating (6) -> good
3. rating (>6)  -> classic 

### Creating histograms for all the input variables and drawing inference 

```{r  echo=FALSE, message =FALSE, warning =FALSE}
p1 =create_hist(ww$fixed.acidity,xname ='fixed.acidity') + 
  scale_x_continuous(limits =c(4,10),breaks = seq(4,10,2))
p2 = create_hist(ww$volatile.acidity,xname ='volatile.acidity') + 
  scale_x_continuous(limits =c(0.0,0.8))
p3 = create_hist(ww$citric.acid,xname ='citric.acid') + 
  xlim(0,1)
p4 = create_hist(ww$residual.sugar,xname ='residual.sugar') + 
  xlim(0,20)
p5 = create_hist(ww$chlorides,xname ='chlorides')+ xlim(0.0,0.2)
p6 = create_hist(ww$free.sulfur.dioxide,xname ='free.sulfur.dioxide') + 
  xlim(0,140)
p7 = create_hist(ww$total.sulfur.dioxide,xname ='total.sulfur.dioxide')
p8 = create_hist(ww$pH,xname ='pH')
p9 = create_hist(ww$sulphates,xname ='sulphates')
p10 = create_hist(ww$alcohol,xname ='alcohol')
p11 = create_hist(ww$density,xname ='density')
# grid,arrange is a function of gridExtra to assemble plots in one frame 
grid.arrange(p1,p2,p3,p4,p5,p6,p7,p8,p9,p10,p11,ncol =3)
```

As the plot suggests for the following variables :

- fixed acidity : shows normal distribution with most of the values between 6-8 (mg / dm^3)
- volatile acidity: shows squed dist with long tail(lots of outliers)
- citric acid: shows normal curve 
- residual sugar : lots of outliers and non - uniformity
- chlorides : normal curve yet lots of outliers
- free sulfur dioxide: squed graph 
- sulphates : bi-modal distribution 
- alcohol : spread unevenly over the whole range

### Scaling uneven graphs to log 10
```{r  echo=FALSE, message =FALSE, warning =FALSE}
 p1 = create_hist(ww$volatile.acidity,xname = 'volatile.acidity (log10)') +
  scale_x_log10() 
# scale_x_log10 scales x axis to log value 
 p2 = create_hist(ww$volatile.acidity,xname = 'volatile.acidity')
 grid.arrange(p2,p1)
```

Above, Using the scale of log to the base 10 the distribution is normalised

```{r  echo=FALSE, message =FALSE, warning =FALSE}
p1 =create_hist(ww$citric.acid, xname = 'citric.acid log10') + 
  scale_x_log10()
p2 = create_hist(ww$citric.acid, xname = 'citric_acid') 
grid.arrange(p2,p1)
```

the tail got shifted from back to front end but noromalising the count histograms of variables are not able to give any insight other than the knowing that 

- the range is widely spread
- there are plenty of outliers 

applying log scale to the sugar residual

```{r  echo=FALSE, message =FALSE, warning =FALSE}
p1 = qplot(residual.sugar, data = ww) + scale_x_log10() + 
  xlab('residual sugar log10')
p2 = qplot(residual.sugar, data = ww) 
grid.arrange(p2,p1)
```

the squed histogram turns out to be bimodal on log scale 

## Exploring relationship between two variables 

Below is a **scatter plot matrix** to look upto for future investigations 

```{r  echo=FALSE, message =FALSE, warning =FALSE,  fig.width=15, fig.height=15}
set.seed(123)
ww_samp <- ww[c(2:13)]
# ggpair function is the function of package GGally which creates the matrix 
ggpairs(ww_samp,lower = list(continuous = wrap("points", shape = I('.'))),
  upper = list(combo = wrap("box", outlier.shape = I('.'))) )

```

the lower half of the diagonals is made of scatter plots of variables in the index and the above is filled with correlation coefficient of
person's correlation.

```{r echo=FALSE, message =FALSE, warning =FALSE}
# function for creating boxplots 
create_box = function(yvariable,y){
  return(qplot(as.factor(quality),yvariable, data = ww, geom = 'boxplot', ylab = y))
}
```


### plotting boxplots for all variables with respect to Quality
```{r  echo=FALSE, message =FALSE, warning =FALSE}
p1 = create_box(ww$citric.acid, y = 'citric.acid')
p2 = create_box(ww$volatile.acidity, y = 'volatile.acidity')
p3 = create_box(ww$fixed.acidity, y = 'fixed.acidity')
p4 = create_box(ww$residual.sugar, y = 'residual.sugar')
p5 = create_box(ww$chlorides, y = 'chlorides')
p6 = create_box(ww$free.sulfur.dioxide, y = 'free.sulfur.dioxide')
p7 = create_box(ww$total.sulfur.dioxide, y = 'total.sulfur.dioxide')
p8 = create_box(ww$pH, y = 'pH')
p9 = create_box(ww$density, y = 'density')
p10 = create_box(ww$alcohol, y = 'alcohol')
p11 = create_box(ww$sulphates, y = 'sulphates')
grid.arrange(p1,p2,p3,p4,p5,p6, ncol =3)
```

```{r  echo=FALSE, message =FALSE, warning =FALSE}
grid.arrange(p7,p8,p9,p10)
```

### Drawing conclusion from the box plots

- lots of outliers in the plots of citric acid, volatile acid and chlorides
- for the quality rating of 9 we can see that there is a distingushed range of quantity
  in all the variables. for example there is a fixed rate of sulfur dioxide for the qaulity 9
  whines other than just one outlier.
- also for density lowest value is recorded against quality rating 9 with one outlier 

### scatter plots between 2 variables 

#### fixed acidity vs density
```{r echo=FALSE, message =FALSE, warning =FALSE}
ggplot(aes(density,fixed.acidity), data = ww) +
  geom_point(alpha = 0.05, position = 'jitter') +
  geom_smooth()

```

Not able to draw any inference out of it . There seems to be a positive slope bu the residual length around it is too
large to be compromised. So i find a dead end to find any link between these two variables

#### total sulfur dioxide vs sulphates
```{r echo=FALSE, message =FALSE, warning =FALSE}
ggplot(aes(sulphates,total.sulfur.dioxide), data = ww) +
  geom_point(alpha = 0.1, position = 'jitter') +
  scale_x_log10()+   # scaling the x axis to log10
  geom_smooth()
```

since the source of sulfur dioxide in wine is sulphate so i chose to plot them against each other. But the plot 
failed to establish the dependancy of sulfur dioxide on sulphates

#### Creating a new variable 
`total_acidity` which adds up `total.acidity` and `volatile.acidity`
```{r  echo=FALSE, message =FALSE, warning =FALSE}
ww$total_acidity =  ww$fixed.acidity + ww$volatile.acidity
summary(ww$total_acidity)
```
Above, summary stats of the variable 'total_acidity'

### plotting total_acidity vs quality
```{r  echo=FALSE, message =FALSE, warning =FALSE}
create_box(ww$total_acidity, y ='total acidity')
```

adding volatile and fixed acidity together and than ploting it against quality.
we can see that average acidity is declining as we should towards better quality.
Only the wines with quality rating 9 are the exception.


**There seems to be a dependency of quality on the variable density as per the boxplot 
lets check out the pearson's coeffecient  between them**

```{r  echo=FALSE, message =FALSE, warning =FALSE}
with(ww, cor.test(density,quality))
```
  
  not a strong coef but shows an acceptable rate of connection between the two
so lets visit the boxplot again to see if there are any trends visible 

```{r  echo=FALSE, message =FALSE, warning =FALSE}
create_box(ww$density, y = 'density')
```
  
  We can see a steady dip in the boxplot range after quality =5 till 9 but  nothing can be
  clearly identified from the scatter plot but wont lose hope and give it another try by 
  plotting mean density with the quality rating values in the form of line graph

```{r  echo=FALSE, message =FALSE, warning =FALSE}
# using dplyr package to chain operations with %>%
ww.qaul_avg = ww %>%  
  group_by(quality)%>%      
summarise(density_avg = mean(density),
          density_median = median(density),
          n =n()) %>%
  arrange(quality)

 ggplot(aes(quality,density_avg),data = ww.qaul_avg) + geom_line() + 
   scale_x_continuous(lim =c(3,10), breaks = seq(3,10,1)) 


```
 
  other than the starting two rating values 3 and 4 the graph shows good linear negative corelation
  
  ## summary of density averaged on the iterations of quality 
```{r echo=FALSE, message =FALSE, warning =FALSE}
summary(ww.qaul_avg)

```
  
  
### Calculating mean values of variables for classic quality wines 
```{r  echo=FALSE, message =FALSE, warning =FALSE}
mean <- c(mean(subset(ww, quality_G == 'classic')$citric.acid),
          mean(subset(ww, quality_G == 'classic')$volatile.acidity),
          mean(subset(ww, quality_G == 'classic')$fixed.acidity),
          mean(subset(ww, quality_G == 'classic')$residual.sugar),
          mean(subset(ww, quality_G == 'classic')$chlorides),
          mean(subset(ww, quality_G == 'classic')$free.sulfur.dioxide),
          mean(subset(ww, quality_G == 'classic')$total.sulfur.dioxide),
          mean(subset(ww, quality_G == 'classic')$density),
          mean(subset(ww, quality_G == 'classic')$pH),
          mean(subset(ww, quality_G == 'classic')$sulphates),
          mean(subset(ww, quality_G == 'classic')$alcohol))


names(mean) <- c('citric.acid', 'volatile.acidity', 'fixed.acidity',
                         'residual.sugar', 'chlorides', 'free.sulfur.dioxide',
                         'total.sulfur.dioxide', 'density', 'pH',
                         'sulphates', 'alcohol')

mean
```


### Calculating median values of variables for classic quality wines 

```{r  echo=FALSE, message =FALSE, warning =FALSE}
median <- c(median(subset(ww, quality_G == 'classic')$citric.acid),
          median(subset(ww, quality_G == 'classic')$volatile.acidity),
          median(subset(ww, quality_G == 'classic')$fixed.acidity),
          median(subset(ww, quality_G == 'classic')$residual.sugar),
          median(subset(ww, quality_G == 'classic')$chlorides),
          median(subset(ww, quality_G == 'classic')$free.sulfur.dioxide),
          median(subset(ww, quality_G == 'classic')$total.sulfur.dioxide),
          median(subset(ww, quality_G == 'classic')$density),
          median(subset(ww, quality_G == 'classic')$pH),
          median(subset(ww, quality_G == 'classic')$sulphates),
          median(subset(ww, quality_G == 'classic')$alcohol))

names(median) <- c('citric.acid', 'volatile.acidity', 'fixed.acidity',
                         'residual.sugar', 'chlorides', 'free.sulfur.dioxide',
                         'total.sulfur.dioxide', 'density', 'pH',
                         'sulphates', 'alcohol')

median
```

## Multi variable exploration 

### fixed acidity vs pH on quality group
```{r  echo=FALSE, message =FALSE, warning =FALSE}
ggplot(aes(pH,fixed.acidity), data = ww) + geom_point(aes(color = ww$quality_G),
                                 alpha =0.1, position = position_jitter(h=0)) + 
  geom_smooth()
# geom_point is for plotting scatterplot 
# geom_smooth  is to smooth out a relation line 
```
```{r  echo=FALSE, message =FALSE, warning =FALSE}
with(ww,cor.test(pH,fixed.acidity))
# cor.test is function to find correlation coeff btw 2 variables 
```
With the graph curve , we can visualise a negetive corelation between the fixed acidity and pH.
scientifically also it makes good sense as more the substance acidic , less is the pH value 
And to back this the co relation factor is also pretty much good to establish a downward slop.


### pH vs total_acidity (coloring with alcohol and faceting with quality_G)
```{r  echo=FALSE, message =FALSE, warning =FALSE}
ggplot(aes(total_acidity,pH), data = ww) + geom_point(aes(color = alcohol),
                          position = position_jitter(h=0) )+ geom_smooth()+
  facet_wrap(~quality_G) 
#facet splits data into different graphs as per the faceted variable 
```

we can observe an inverse relation with the smooth line between the acidity and pH which can be understood 
scientifically as more the acidity lesser the pH on a scale of 1 to 14

### residual.sugar vs density (coloring with alcohol)
```{r  echo=FALSE, message =FALSE, warning =FALSE}
 ggplot(aes(density,residual.sugar),
        data = subset(ww,quantile(residual.sugar)<0.99)) + 
  geom_point(aes(color = alcohol),position = position_jitter(h=0)) +
  scale_x_continuous(lim =c(0.98,1.04), breaks = seq(0.98,1.04,0.02) )
 
```

a linear relationship can be esatblished from this graph between residual sugar and density.


### alcohol vs total.sulfur.dioxide (color = as.factor(quality))

```{r echo=FALSE, message =FALSE, warning =FALSE}
set.seed(123)   # setting seed to recreate same sample set every time
ww_sample = ww[sample(1:length(ww$quality),1000),] #sampled with 1000 obs
ggplot(aes(total.sulfur.dioxide,alcohol), data = ww_sample) + 
  geom_point(aes(color = as.factor(quality)), position = position_jitter(h=0)) +
  geom_smooth(color = 'black') +
  scale_color_brewer(type = 'seq')
```

better quality wines are observed on the upper side of the graph, closer to the y axis.

# Final plots and summary

## Plot 1 : alcohol vs total.sulfur.dioxide

```{r echo=FALSE, message =FALSE, warning =FALSE}
set.seed(123)   # setting seed to recreate same sample set every time
ww_sample = ww[sample(1:length(ww$quality),1000),] #sampled with 1000 obs
ggplot(aes(total.sulfur.dioxide,alcohol), data = ww_sample) + 
geom_point(aes(color = as.factor(quality))) + 
scale_color_brewer(type = 'seq', 
                   palette = 'Blues',
                   guide = guide_legend
                   (title = 'quality', 
                   reverse = T,
                   override.aes = list(alpha = 1, size = 2))) +   
  # color palatte , naming the legend, fixing order, size , alpha 
geom_smooth(color = 'black') +
scale_x_continuous(lim = c(0,300), breaks = seq(0,300,20)) + # scaling axis 
scale_y_continuous(lim = c(8,14), breaks = seq(8,14,0.5)) +
ggtitle('Alcohol vs Total sulfur dioxide on wine quality') +  # naming plot 
ylab('alcohol (% by volume)') +           # naming axis
xlab('Total sulfur dioxide (mg / dm^3)')  
  
```

## Inference of plot 1
From this graph we can conclude that alcohol has negative correltion with sulfur dioxide and we can see that **wines of rating above 6**
are mostly placed in a region of ** more alcohol % by volume and less of sulfur dioxide**

## Plot 2
```{r echo=FALSE, message =FALSE, warning =FALSE}
     ggplot(aes(density,residual.sugar),
            data = subset(ww,quantile(residual.sugar)<0.99)) + 
  # taking lesser than 99 per centil^
  geom_point(aes(color = alcohol), 
             position = position_jitter(h=0), size =1 ) + # scatter plot
  facet_wrap(~quality_G) + # faceting as per quality group
  scale_y_continuous(lim = c(0,20), breaks = seq(0,20,5)) + # scaling y axis 
  ggtitle('Residual sugar vs Density on Alcohol faceted by Quality Group') +
  xlab('Density (g / cm^3)') +      # x axis label
  ylab('Residual sugar (g / dm^3)') # y axis label 


```

##  Inference of plot 2
what we can conclude from this plot is that there is a  strong linear relationship between the residual sugar and 
density . The important point over here is that ** at the bottom of the graph closer to 0 axis alcohol per centage is higher** which points out the fact that **lesser the sugar and density more is the alcohol %** and **more the alcohol better is the wine qaulity**. So we can indirectly link sugar and density with quality of wine on an inverse relation.


## Plot 3
```{r echo=FALSE, message =FALSE, warning =FALSE}

     ggplot(data = ww, aes(as.factor(quality),   
                           total.sulfur.dioxide)) +
               geom_boxplot() +     # lplotting boxplot
               xlab("Quality") +    # lableing x axis
               ylab("Total Sulfur Dioxide (g / dm^3)") + # labeling x axis
              ggtitle("Total Sulfur Dioxide vs Quality Boxplot")  # title 

```

##  Inference of plot 3
boxplot : we can see that as quality of wine increases the deviation of value of sulfur dioxide from the median 
value decreases closing on to an optimal value for quality improvement.

# Reflection 
So after going through all of the EDA process i was able to connect most of the input variables to the outcome quality 
variable. To draw a final conclusion a higher level of alcohol % v/v , less possible density and residual sugar and an 
optimal value of total sulfur dioxide is the formula to create top quality wine.

Mostly the path i used to follow was strongly driven by scatter plot matrix.

The place where i got stuck was coming out with the third final plot to back a good hypothysis. The 3rd plot i came up with
took me time to trace out. 

Again the conclusion strongly relies on the quality ratings given by the critics which can differ as per taste or bias to 
a particular brand.

What future work can be done is to build a model to rate the wines from this data set and to find a configuration of elements to make top quality wine , the process should be focused on the chemistry of substances as well doing tests in the labs and 
finding out a perfect wine by trial and error method in labs and bruery.

###ThankYou

